---
title: "YRBS_Analysis"
author: "Michael Perez, Seth Galluci, Brian Bellamy, Harold Gonzalez"
date: "2025-12-02"
output: 
  html_document:
  toc: true
  toc_float: true
  number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Install packages as needed with install.packages()
library(survey)
library(haven)
library(here)
library(ggplot2)
library(tidyr)
library(tidyverse)
library(readr)
library(dplyr)
library(broom)
library(viridis)
library(survey)
library(patchwork)
library(broom)
library(labelled)
```


```{r read dataset, echo=FALSE}
yrbs23 <- read_sav(here("2023 Data/yrbs2023.sav"))

#Recoding strips the labels sometimes, in order to avoid it, I am pre-saving the labels, and reapplying them after recoding.
original_labels <- lapply(yrbs23, function(x)
  attr(x, "label"))
original_value_labels <- lapply(yrbs23, function(x)
  attr(x, "labels"))
```

#Data Cleaning

Ensured ordinals are all worse = higher, flipped binaries from 1-2 to 0-1 to match. 

```{r , echo=FALSE}
#Identifies all binary variables in the dataset
binary_vars <- names(Filter(function(x) {
  vals <- unique(x)
  all(vals %in% c(1, 2, NA))
}, yrbs23))

#Recodes all of the 2's to a 0 in binary variables so that graphs don't show up upside-down. Yes is almost always the worse answer in these questions, so Yes should be on top.
yrbs23[binary_vars] <- lapply(yrbs23[binary_vars], function(x) {
  ifelse(is.na(x), NA, ifelse(x == 1, 1, 0))
})

#If we run into any that we analyze that the No should be on top, we can individually recode that one back to a 2

#In Q87, 6 and 7 are "None of these" and "Don't Know". This throws off the scale, so they are getting N/A'd.
yrbs23$Q87 <- ifelse(yrbs23$Q87 %in% c(6, 7), NA, yrbs23$Q87)

#Recoding Q80 - 2 and 3 mean the same thing essentially "A few time a month" "Once a week", and they have much fewer responses than the other categories, which may be spiking the graphs and screwing up the regression curves.

yrbs23$Q80old <- yrbs23$Q80

yrbs23$Q80 <- case_when(
  yrbs23$Q80 == 1 ~ 1,
  yrbs23$Q80 %in% c(2, 3) ~ 2,
  yrbs23$Q80 == 4 ~ 3,
  yrbs23$Q80 == 5 ~ 4,
  yrbs23$Q80 == 6 ~ 5,
  yrbs23$Q80 == 7 ~ 6,
  yrbs23$Q80 == 8 ~ 7,
  TRUE ~ NA_real_
)


#Removes labels for the codes that no longer exist
attr(yrbs23$Q87, "labels") <- attr(yrbs23$Q87, "labels")[1:5]

#Reapplying labels
for (v in binary_vars) {
  if (!is.null(original_labels[[v]])) {
    var_label(yrbs23[[v]]) <- original_labels[[v]]
  }
}

for (v in binary_vars) {
  val_labels(yrbs23[[v]]) <- c("No" = 0, "Yes" = 1)
}

# Restore Q87 variable label
var_label(yrbs23$Q87) <- original_labels[["Q87"]]

orig_vals_Q87 <- original_value_labels[["Q87"]]
clean_vals_Q87 <- orig_vals_Q87[names(orig_vals_Q87) %in% c("1", "2", "3", "4", "5")]

val_labels(yrbs23$Q87) <- clean_vals_Q87
table(yrbs23$Q80, useNA = "ifany")

```
```{r, echo=FALSE}
#Survey weighting. Regression models need to use this.

yrbsdes23 <- svydesign(
  id     = ~ psu,
  strata = ~ stratum,
  weights = ~ weight,
  data   = yrbs23,
  nest   = TRUE
)

```

```{r, echo=FALSE}
#Individual simple regression loop for exploratory analysis

table(yrbs23$Q80, useNA = "ifany")


sig_vars <- c("Q42",
              "Q33",
              "Q36",
              "Q93",
              "Q56",
              "Q16",
              "Q12",
              "Q24",
              "Q26",
              "Q27",
              "Q84",
              "Q87")

results <- data.frame(
  Variable = character(),
  Level = character(),
  Estimate = numeric(),
  P_Value = numeric(),
  stringsAsFactors = FALSE
)

for (v in sig_vars) {
  #
  fmla <- as.formula(paste("Q80 ~", v))
  
  # Fit model
  m <- svyglm(fmla, design = yrbsdes23)
  
  # Tidy output
  tidy_m <- tidy(m)
  
  # Keep only the rows that are NOT the intercept
  effects <- tidy_m[tidy_m$term != "(Intercept)", ]
  
  # Add each effect to results table
  for (i in seq_len(nrow(effects))) {
    results <- rbind(
      results,
      data.frame(
        Variable = v,
        Level = effects$term[i],
        Estimate = effects$estimate[i],
        P_Value = effects$p.value[i]
      )
    )
  }
}

results
```

```{r, echo=FALSE}
#Notice, I set eval=FALSE on this block. It will be skipped in a knit and a run-all. If you want this printout, you need to run this block individually. I was trying to keep your computer from exploding.

all_Q_vars <- paste0("Q", 1:107)
present_Q <- intersect(all_Q_vars, names(yrbs23))

# Keep only numeric Q variables with >1 unique value
is_num <- sapply(present_Q, function(v) {
  x <- yrbs23[[v]]
  is.numeric(x) || inherits(x, "haven_labelled")
})

has_variation <- sapply(present_Q, function(v) {
  x <- yrbs23[[v]]
  length(unique(x[!is.na(x)])) > 1
})

numeric_Q_vars <- present_Q[is_num & has_variation]

# Don’t regress Q80 on itself
sig_vars <- setdiff(numeric_Q_vars, "Q80")

results <- data.frame(
  Variable = character(),
  Level = character(),
  Estimate = numeric(),
  P_Value = numeric(),
  stringsAsFactors = FALSE
)

for (v in sig_vars) {
  fmla <- as.formula(paste(v, "~ Q80"))
  m <- try(svyglm(fmla, design = yrbsdes23), silent = TRUE)
  
  # Skip if model failed
  if (inherits(m, "try-error"))
    next
  
  tidy_m <- tidy(m)
  effects <- tidy_m[tidy_m$term != "(Intercept)", ]
  
  if (nrow(effects) > 0) {
    results <- rbind(
      results,
      data.frame(
        Variable = v,
        Level    = effects$term,
        Estimate = effects$estimate,
        P_Value  = effects$p.value,
        stringsAsFactors = FALSE
      )
    )
  }
}

results
```

```{r variable check and some setup, echo=FALSE}
# Just checking to make sure the variables are coded as expected

# This runs a print that shows the variable name, label, value coding, and frequencies for a sanity check.

table(yrbs23$Q42, useNA = "ifany")

attr(yrbs23$Q42, "labels")

#Really, feel free to add any you're interested in here. As we "disqualify" some of these by proving there is no interesting correlation, we can focus on others.
vars_to_check <- c("Q12",
                   "Q16",
                   "Q24",
                   "Q26",
                   "Q27",
                   "Q33",
                   "Q36",
                   "Q42",
                   "Q56",
                   "Q84",
                   "Q87",
                   "Q93")

#Since we're at defining groups of vars to use for various purposes, here's a couple more reusable variables.

#Variables in the dataset, that we care about, that use a 0-1 scale (after recoding)
bin_vars <- c("Q24", "Q26", "Q27", "Q56")

#Variables in the dataset, that we care about, that use a 1-7 scale
ord_vars <- c("Q12", "Q16", "Q33", "Q36", "Q42", "Q84", "Q87", "Q93")

#Used later
all_plot_vars <- setdiff(c(bin_vars, ord_vars), "Q80")



#A neat piece of copyable code also, use this to make good labels. It pulls the "real" column name from the Q designator.
#paste(attr(yrbs23[[v]], "label"))

for (v in vars_to_check) {
  cat("\n==============================\n")
  cat("Variable:", v, "\n")
  
  # Print SPSS variable label (the descriptive text)
  var_label <- attr(yrbs23[[v]], "label")
  cat("Label:", var_label, "\n")
  
  # Print value labels (1=..., 2=..., etc.)
  value_labels <- attr(yrbs23[[v]], "labels")
  cat("Value Labels:\n")
  print(value_labels)
  
  # Print observed frequencies
  cat("Frequencies:\n")
  print(table(yrbs23[[v]], useNA = "ifany"))
}

```

Plotting

```{r, echo=FALSE}
#Multi box-plot of all variables.
plot_list <- lapply(all_plot_vars, function(v) {
  ggplot(yrbs23 %>% drop_na(all_of(v), Q80), aes(x = factor(Q80), y = .data[[v]])) +
    geom_boxplot() +
    labs(x = "Social Media Use (Q80)", y = v, title = paste(attr(yrbs23[[v]], "label")))
})

wrap_plots(plotlist = plot_list, ncol = 3)
```


```{r, echo=FALSE}
#Ensures the proportional barplots treat binaries and ordinals correctly.
is_binary <- function(v) {
  all(unique(yrbs23[[v]][!is.na(yrbs23[[v]])]) %in% c(0, 1))
}

#Proportional barplot
plot_list <- lapply(all_plot_vars, function(v) {
  df <- yrbs23 %>% drop_na(Q80, all_of(v))
  
  if (is_binary(v)) {
    df <- df %>% mutate(any_behavior = .data[[v]])
  } else {
    df <- df %>% mutate(any_behavior = ifelse(.data[[v]] == 1, 0, 1))
  }
  
  temp <- df %>% group_by(Q80) %>% summarise(percent = mean(any_behavior) * 100)
  
  ggplot(temp, aes(x = factor(Q80), y = percent)) +
    geom_col(fill = "steelblue", alpha = 0.8) +
    geom_text(aes(label = sprintf("%.1f%%", percent)), vjust = -0.5) +
    labs(x = "Social Media Use (Q80)", y = "Percent (%)", title = paste0(v, ": ", attr(yrbs23[[v]], "label"))) +
    ylim(0, 100) +
    theme_minimal()
})

wrap_plots(plotlist = plot_list, ncol = 3)
```

```{r stacked bars, echo=FALSE}
make_stacked_bar <- function(v) {
  yrbs23 %>%
    drop_na(Q80, all_of(v)) %>%
    mutate(Q80 = factor(Q80), outcome = factor(.data[[v]])) %>%
    count(Q80, outcome) %>%
    group_by(Q80) %>%
    mutate(prop = n / sum(n)) %>%
    ggplot(aes(x = Q80, y = prop, fill = outcome)) +
    geom_col(color = "black", linewidth = 0.1) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(
      title = paste0(v, ": ", attr(yrbs23[[v]], "label")),
      x = "Social Media Use (Q80)",
      y = "Percent",
      fill = v
    )  +
    theme_minimal(base_size = 11)
}

stacked_plots <- lapply(all_plot_vars, function(v) {
  make_stacked_bar(v)
})

wrap_plots(plotlist = stacked_plots, ncol = 3)

```


```{r, echo=FALSE}
make_mean_plot <- function(v) {
  yrbs23 %>%
    drop_na(Q80, all_of(v)) %>%
    mutate(Q80_num = as.numeric(Q80)) %>%
    group_by(Q80_num) %>%
    summarise(mean_value = mean(.data[[v]])) %>%
    ggplot(aes(x = Q80_num, y = mean_value)) +
    geom_line(size = 1.1, color = "steelblue") +
    geom_point(size = 3, color = "steelblue") +
    scale_x_continuous(breaks = sort(unique(as.numeric(yrbs23$Q80)))) +
    labs(
      title = paste0(v, ": ", attr(yrbs23[[v]], "label")),
      x = "Social Media Use (Q80)",
      y = paste("Mean", v, "(higher = worse)")
    ) +
    theme_minimal(base_size = 11)
}


mean_plots <- lapply(all_plot_vars, make_mean_plot)
wrap_plots(plotlist = mean_plots, ncol = 3)
```
```{r}
current_Q_vars <- paste0("Q", 1:12)
mean_plots <- lapply(current_Q_vars, make_mean_plot)
wrap_plots(plotlist = mean_plots, ncol = 3)
```
```{r}
stacked_plots <- lapply(current_Q_vars, function(v) {
  make_stacked_bar(v)
})

wrap_plots(plotlist = stacked_plots, ncol = 3)
```
```{r}
current_Q_vars <- paste0("Q", 13:24)
mean_plots <- lapply(current_Q_vars, make_mean_plot)
wrap_plots(plotlist = mean_plots, ncol = 3)
```
```{r}
stacked_plots <- lapply(current_Q_vars, function(v) {
  make_stacked_bar(v)
})

wrap_plots(plotlist = stacked_plots, ncol = 3)
```
```{r}
current_Q_vars <- paste0("Q", 25:36)
mean_plots <- lapply(current_Q_vars, make_mean_plot)
wrap_plots(plotlist = mean_plots, ncol = 3)
```
```{r}
stacked_plots <- lapply(current_Q_vars, function(v) {
  make_stacked_bar(v)
})

wrap_plots(plotlist = stacked_plots, ncol = 3)
```
```{r}
current_Q_vars <- paste0("Q", 37:48)
mean_plots <- lapply(current_Q_vars, make_mean_plot)
wrap_plots(plotlist = mean_plots, ncol = 3)
```
```{r}
stacked_plots <- lapply(current_Q_vars, function(v) {
  make_stacked_bar(v)
})

wrap_plots(plotlist = stacked_plots, ncol = 3)
```
```{r}
current_Q_vars <- paste0("Q", 49:62)
mean_plots <- lapply(current_Q_vars, make_mean_plot)
wrap_plots(plotlist = mean_plots, ncol = 3)
```
```{r}
stacked_plots <- lapply(current_Q_vars, function(v) {
  make_stacked_bar(v)
})

wrap_plots(plotlist = stacked_plots, ncol = 3)
```
```{r}
current_Q_vars <- paste0("Q", 63:74)
mean_plots <- lapply(current_Q_vars, make_mean_plot)
wrap_plots(plotlist = mean_plots, ncol = 3)
```
```{r}
stacked_plots <- lapply(current_Q_vars, function(v) {
  make_stacked_bar(v)
})

wrap_plots(plotlist = stacked_plots, ncol = 3)
```
```{r}
current_Q_vars <- paste0("Q", 75:86)
mean_plots <- lapply(current_Q_vars, make_mean_plot)
wrap_plots(plotlist = mean_plots, ncol = 3)
```
```{r}
stacked_plots <- lapply(current_Q_vars, function(v) {
  make_stacked_bar(v)
})

wrap_plots(plotlist = stacked_plots, ncol = 3)
```
```{r}
current_Q_vars <- paste0("Q", 87:98)
mean_plots <- lapply(current_Q_vars, make_mean_plot)
wrap_plots(plotlist = mean_plots, ncol = 3)
```
```{r}
stacked_plots <- lapply(current_Q_vars, function(v) {
  make_stacked_bar(v)
})

wrap_plots(plotlist = stacked_plots, ncol = 3)
```
```{r}
current_Q_vars <- paste0("Q", 99:107)
mean_plots <- lapply(current_Q_vars, make_mean_plot)
wrap_plots(plotlist = mean_plots, ncol = 3)
```
```{r}
stacked_plots <- lapply(current_Q_vars, function(v) {
  make_stacked_bar(v)
})

wrap_plots(plotlist = stacked_plots, ncol = 3)
```
This *MIGHT* be enough graphs. And we might not even need them all. Or maybe we do more. I was trying to avoid getting my brain stuck in regression since we've been talking about it in class for so long. 

We should go back to the beginning of the semester and look back at some of the ways we described statistics, some of those might be more helpful for these types of problems, since when we do have slopes, they aren't always very strong. 

We also need to sometimes look at these from multiple angles, for example, in the alcohol chart, the correlations get a lot more clear when I recoded it to a binary and represented it proportionally. 

A good approach might be to demonstrate which variables do/don't have relationships with social media using the charts, then use the regression data to define and describe those relationships. 

---